Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-02-14 Dail_Mad_ext.tex
Several minor edits

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in 
bias, and thus ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential, Ricker-logistic, etc $\dots$), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the \textbf{R} package
\texttt{unmarked} or \textbf{JAGS} is also provided.
A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested the
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is unavailable.
}

@@

In this paper, we focus on the model of \citet[henceforth the DM model;]{dail_madsen:2011}
that simultaneous resolves each of the aforementioned problems with
traditional state-space models, and is designed for simple countdata.
In the following section, we describe the DM model in its original
form and explain how it resolves each of the deficiencies with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time-series. Specifically, we describe methods for accommodating
excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of the paper is to provide ecologists with flexible and accessible means of
addressing important questions related to the variation of abundance in
space and time.
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.
\begin{comment}
  The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error
using only spatially and temporally replicated count data. Their model
is an extension of the closed-population $N$-mixture model
\citep{royle:2004biom}, which was designed specifically to
address the problem of modeling spatial variation in abundance in the
  presence of observation error, specifically, when detection
  probability is less than one. The DM extension relaxes the
assumption of population closure and includes explicit
parameters describing population change as a first-order Markovian
  process. [more details]

@@

The DM model requires count data collected at $R$ sites, each of
which is surveyed on $T$ primary sampling periods. A site is ideally a
well-defined region of a study area, such as a wetland or a
patch of early-successional habitat, although it could be an arbitrarily
defined region such as a randomly located survey plots. The time
frame of the study is arbitrary, but in general
it will be sufficiently long to allow for temporal variation in
abundance. For instance, a primary sampling period could be a
one month long breeding season, which could be surveyed once per
year for $T$ years. If the a site is surveyed on $J$ occasions during a
primary period, these are called secondary sampling occasions.
In the case where no secondary sampling was conducted, i.e.
$J=1$, let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
at site $i$ and primary period $t$. If secondary sampling periods were
used, an extra dimension is added so that we have $X_{ijt}:
j=1,\hdots,J$. Typically, we $X_{it}$ will be lower than the
actual quantity of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $X_{it} \equiv N_{it}$.

@@

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
In the Poisson case, the equivalent expression of
$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
\begin{equation}
[N_{i1} | \Lambda] =
\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
  \label{eq:N1br}
\end{equation}

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
i.e. the model for abundance dynamics. This state
model has allowed us to replace the initial abundance
distribution and transition models of conventional state-space models, with
alternatives that (1) allow for modeling spatial variation in
abundance, (2) respect the discrete nature of the counts, and
(3) have parameters that are estimable. The one
remaining task is to replace the observation model
$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection, and the
one considered in the original formulation, is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
or equivalently,
\begin{equation}
[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
  \label{eq:1}
\end{equation}
Under this model, $\bm{p}$ are interpretted as detection
probability
parameters...

@@

Writeing custom MCMC algorthims is often tedious and foreign to
ecologists, but software packages such as \textbf{WinBUGS} and
\textbf{JAGS} overcome the technical difficulties by allowing
users to specify the model using a simple symbolic descriptions. For
example here is \textbf{BUGS} code for the ``geometric'' model described
shown
in Eq.~\ref{eq:hm}:
\begin{verbatim}
model {
lambda ~ dunif(0, 5)
omega ~ dunif(0,1)
gamma ~ dunif(0, 10)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    S[i,t-1] ~ dbin(omega, N[i,t-1])
    G[i,t-1] ~ dpois(gamma*N[i,t-1])
    N[i,t] <- S[i,t-1] + G[i,t-1]
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}

@@

%% Is this here on purpose?  We haven't introduced it, and I was referring to 
%% the speed-up these models provide in the maximum-likelihood framework
%% (even the classic DM models don't require the convolution in the Bayesian framework, right?).
\begin{verbatim}
model {
lambda ~ dunif(0, 5)
r ~ dunif(-5, 5)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    mu[i,t-1] = N[i,t-1]*exp(r)
    N[i,t] ~ dpois(mu[i,t-1])
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}

@@

Estimates for the top ranked model for ovenbird
%(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota(.)$]p(wind+1st)) 
were similar when
run in the Bayesian framework, except that the estimate of r
more than halved (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for
Bayesian model estimates we present mean and SD]). When random observer effects
were added, estimates for $\Lambda$ and K increased dramatically
(from 31.6 $\pm$ 4.9 to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1,
respectively), and the estimate for r was intermediate (0.018 $\pm$ 0.006). The
estimate of the intercept for p (on the logit scale) dropped from -1.5 $\pm$
0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.
Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
provided some evidence for a lack of convergence in estimates of the
intercept for p for all Bayesian models ran.

@@

Estimates for the top ranked model for golden-winged warbler
%(ZIP[Λ(.)α(.)]Exponential[r(.)]p(1st))
were similar when run in the Bayesian framework. When random
observer effects were added, the estimate for $\Lambda$ increased dramatically
(from 22.4 $\pm$ 7.6 to 55.5 $\pm$ 30.9). The estimate of the intercept
for $p$ (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. Adding zero-inflation to the dynamics (as well as
the initial abundance) has no effect for exponential models, because populations
that started at 0 would stay at 0 in any case. Therefore, we tested
zero-inflated dynamics on the Exponential + Immigration model. Adding zero-inflation to the dynamics
increased the estimate of ι from 0 to 0.091 $\pm$ 0.049, but had only minor
effects on the other parameters. Gelman and Rubin diagnostics and visual
examination of the chain trajectory and density plots provided evidence for a lack of convergence in
estimates of $\Lambda$ for all Bayesian models ran, and for most parameters when random
effects were included.

@@

Even though the DM models and our extentions are conceptually simple, even in
their most stripped down form they include many random effects (the $N$'s, which
incorporate demographic stochasticity), making it challenging to compute the likelihood 
or implement MCMC algorithms. Nonetheless, in some cases additional random effects
may be of interest.  We have shown a few approaches for incorporating some of these
(environmental stochasticity and random observer effects).  Other approaches to modeling
each of these may also be worth trying.  Incorporating these random effects can increase
the difficulty of achieving model convergence, but they are certainly important.
[Something about how much including each of these random effects changes ovenbird estimates.]

>>>>

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in 
bias, and thus ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the \textbf{R} package
\texttt{unmarked} or \textbf{JAGS} is also provided.
A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested the
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is unavailable.
}

@@

In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
that simultaneous resolves each of the aforementioned problems with
traditional state-space models, and is designed for simple count data.
In the following section, we describe the DM model in its original
form and explain how it resolves each of the deficiencies with
standard state-space models. In Section~\ref{sec:ext}, we extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time-series. Specifically, we describe methods for accommodating
excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of the paper is to provide ecologists with flexible and accessible means of
addressing important questions related to the variation of abundance in
space and time.
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.
\begin{comment}
  The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error
using only spatially and temporally replicated count data. Their model
is an extension of the closed-population $N$-mixture model
\citep{royle:2004biom}, which was designed specifically to
address the problem of modeling spatial variation in abundance in the
  presence of observation error, specifically, when detection
  probability is less than one. The DM extension relaxes the
assumption of population closure and includes explicit
parameters describing population change as a first-order Markovian
  process. [more details]

@@

The DM model requires count data collected at $R$ sites, each of
which is surveyed on $T$ primary sampling periods. A site is ideally a
well-defined region of a study area, such as a wetland or a
patch of early-successional habitat, although it could be an arbitrarily
defined region such as a randomly located survey plots. The time
frame of the study is arbitrary, but in general
it will be sufficiently long to allow for temporal variation in
abundance. For instance, a primary sampling period could be a
one month long breeding season, which could be surveyed once per
year for $T$ years. If the a site is surveyed on $J$ occasions during a
primary period, these are called secondary sampling occasions.
In the case where no secondary sampling was conducted, i.e.
$J=1$, let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
at site $i$ and primary period $t$. If secondary sampling periods were
used, an extra dimension is added so that we have $X_{ijt}:
j=1,\hdots,J$. Typically, we know $X_{it}$ will be lower than the
actual quantity of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $X_{it} \equiv N_{it}$.

@@

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
In the Poisson case, the equivalent expression of
$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
\begin{equation}
[N_{i1} | \Lambda] =
\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
  \label{eq:N1br}
\end{equation}
% Maybe we don't need the Poisson distriubtion equation? - Jeff

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
i.e. the model for abundance dynamics. This state
model has allowed us to replace the initial abundance
distribution and transition models of conventional state-space models, with
alternatives that (1) allow for modeling spatial variation in
abundance, (2) respect the discrete nature of the counts, and
(3) have parameters that are estimable. The one
remaining task is to replace the observation model
$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
or equivalently,
\begin{equation}
[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
  \label{eq:1}
\end{equation}
Under this model, $\bm{p}$ are interpreted as detection
probability parameters.

@@

Writeing custom MCMC algorthims is often tedious and foreign to
ecologists, but software packages such as \textbf{WinBUGS} and
\textbf{JAGS} overcome the technical difficulties by allowing
users to specify the model using a simple symbolic descriptions. For
example here is \textbf{BUGS} code for the ``geometric'' model described
in Eq.~\ref{eq:hm}:
\begin{verbatim}
model {
lambda ~ dunif(0, 5)
omega ~ dunif(0,1)
gamma ~ dunif(0, 10)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    S[i,t-1] ~ dbin(omega, N[i,t-1])
    G[i,t-1] ~ dpois(gamma*N[i,t-1])
    N[i,t] <- S[i,t-1] + G[i,t-1]
    n[i,t] ~ dbin(p, N[i,t])
    }
  }
}
\end{verbatim}

@@

@@

Estimates for the top ranked model for ovenbird
%(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota(.)$]p(wind+1st)) 
were similar when
run in the Bayesian framework, except that the estimate of $r$
more than halved (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for
Bayesian model estimates we present mean and SD]). When random observer effects
were added, estimates for $\Lambda$ and $K$ increased dramatically
(from 31.6 $\pm$ 4.9 to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1,
respectively), and the estimate for r was intermediate (0.018 $\pm$ 0.006). The
estimate of the intercept for $p$ (on the logit scale) dropped from -1.5 $\pm$
0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  When
environmental stochasticity was added, the estimate for $K$ increased (to 112 $\pm$ 58.8) and $r$ was
again intermediate (0.020 $\pm$ 0.010).  Adding environmental stochasticity alone had little effect
on estimate of $p$ or $\Lambda$. The highest estimates of $\Lambda$ and $K$ came 
from the model that included environmental stochasticity and random observer effects (43.8 $\pm$
7.9 and 141.5 $\pm$ 69.5, respectively).  
Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
provided some evidence for a lack of convergence in estimates of the
intercept for $p$ for all Bayesian models ran.

@@

Estimates for the top ranked model for golden-winged warbler
%(ZIP[Λ(.)α(.)]Exponential[r(.)]p(1st))
were similar when run in the Bayesian framework. When random
observer effects were added, the estimate for $\Lambda$ increased dramatically
(from 22.4 $\pm$ 7.6 to 55.5 $\pm$ 30.9). The estimate of the intercept
for $p$ (on the logit scale) dropped from -3.5 $\pm$ 0.2 to -5.6 $\pm$ 1.1, and the
estimate of $\sigma_p$ was 1.7 $\pm$ 0.7. Adding environmental stochasticity
also increased the estimate of $\Lambda$ (to 42.9 $\pm$ 17.9) and the model with
both random effects had the highest estimate of $\Lambda$ (74.6 $\pm$ 32.9). 
Adding zero-inflation to the dynamics (as well as
the initial abundance) has no effect for exponential models, because populations
that started at 0 would stay at 0 in any case. Therefore, we tested
zero-inflated dynamics on the Exponential + Immigration model. Adding zero-inflation to the dynamics
increased the estimate of $\iota$ from 0 to 0.091 $\pm$ 0.049, but had only minor
effects on the other parameters. Gelman and Rubin diagnostics and visual
examination of the chain trajectory and density plots provided evidence for a lack of convergence in
estimates of $\Lambda$ for all Bayesian models ran, and for most parameters when random observer
effects were included.

@@

Even though the DM models and our extentions are conceptually simple, even in
their most stripped down form they include many random effects (the $N$'s, which
incorporate demographic stochasticity), making it challenging to compute the likelihood 
or implement MCMC algorithms. Nonetheless, in some cases additional random effects
may be of interest.  We have shown a few approaches for incorporating some of these
(environmental stochasticity and random observer effects).  Other approaches to modeling
each of these may also be worth trying.  Incorporating these random effects can increase
the difficulty of achieving model convergence, but they are certainly important.
For example, the estimate of carrying capacity for the ovenbird increased by more than 150\%
when environmental stochasticity and random observer effects were included.
